---
title: "Section 4: Statistical Relationships"
author: "Sam Frederick"
institute: "Columbia University"
format: 
  revealjs:
    css: [footer.css]
    theme: [default, custom.scss]
    chalkboard: true
    progress: true
    include-after-body: footer.html 
    footer: "Statistical Relationships"
knitr: 
  opts_chunk: 
      echo: true
---

## Today's Section

- Correlation Coefficients

- Law of Large Numbers

- The Central Limit Theorem

- Hypothesis Testing

## Correlation Coefficients

:::{.fragment}
- Statistically summarize relationships between numeric variables
:::
:::{.fragment}
- Range from -1 to 1
:::
:::{.fragment}
- Values closer to -1 or 1 indicate stronger relationships
:::

## Correlation Coefficients

$Cor(X, Y) = \frac{Cov(X,Y)}{\sqrt{Var(X)*Var(Y)}} = \frac{\frac{1}{n-1}\sum_{i=1}^n(X-\bar{X})(Y-\bar{Y})}{\sqrt{\frac{1}{n-1}\sum_{i=1}^n(X-\bar{X})^2 *\frac{1}{n-1} \sum_{i=1}^n(Y-\bar{Y})^2}}$

Equivalently:

$Cor(X,Y) = \frac{\sum_{i=1}^n (X-\bar{X})*(Y-\bar{Y})}{ \sqrt{\sum_{i=1}^n(X-\bar{X})^2 *\sum_{i=1}^n(Y-\bar{Y})^2}}$

## Correlation Coefficients in R

```{r}
x <- c(1, 3, 5, 7, 9)
y <- c(1, 4, 7, 2 , 10)
```

:::{.fragment}
- By hand: 

```{r}
numerator <- sum((x-mean(x))*(y - mean(y)))
denominator <- sqrt(sum((x-mean(x))^2)*sum((y - mean(y))^2))
round(numerator/denominator, 3)
```
:::
:::{.fragment}
- Using R functions:

```{r}
round(cov(x, y)/sqrt(var(x)*var(y)), 3)
```
:::
:::{.fragment}
- Using one R function:

```{r}
round(cor(x, y), 3)
```
:::

## Law of Large Numbers

As $n \to \infty$, sample mean approaches true population mean:

```{r, echo = F}
library(tidyverse)
set.seed(123)
pop <- rnorm(1000000, 375, 1000)
samp_sizes <- seq(1, 10000, 5)
samp_means <- sapply(samp_sizes, 
       function(x) mean(sample(pop, size = x, replace = T)))
tibble(samp_mean = samp_means, 
       samp_size = samp_sizes) %>%
  ggplot(aes(samp_size, samp_mean)) + geom_line() +
  geom_hline(yintercept = mean(pop), col = "red", lty = "dashed") +
  labs(x = "Sample Size", y = "Sample Mean", title = "Law of Large Numbers") 
```

## Central Limit Theorem

- As $n \to \infty$, $\bar{X}_n \to N(\mu, \frac{\sigma^2}{\sqrt{n}})$

```{r, echo = F}
set.seed(12345)
pop <- sample(1:30, 1000000, replace = T)
clt <- lapply(c(1:30, 50,75, 100,250, 500, 
                1000,2500,  5000, 7500, 10000, 100000), 
       function(x) replicate(200, mean(sample(pop, size = x, replace = T))))

```

## Central Limit Theorem

- What's more,  As $n \to \infty$, $\frac{\bar{X}_n - \mu}{\sigma/\sqrt{n}} \to N(0, 1)$

```{r, echo = F, cache = T}
a <- tibble(sample = unlist(clt),
       sampsize = rep(c(1:30, 50,75, 100,250, 500, 1000,2500,  5000, 7500, 10000, 100000), lengths(clt))) %>%
  group_by(sampsize) %>%
  mutate(sd_samp = sd(pop)/sqrt(sampsize)) %>%
  ungroup() %>%
  mutate(sample =(sample - mean(pop))/(sd_samp), 
         sampsize = factor(sampsize)) %>% 
  ggplot(aes(sample)) + 
    geom_histogram(aes(y = ..density..)) +
    stat_function(fun = dnorm, color = "red") + 
    labs(x = "Sample Means", y = "Density", title = "Sample Size: {closest_state}")+
  # facet_wrap(~sampsize)
    gganimate::transition_states(sampsize,transition_length = 0)
gganimate::animate(a, duration = 10)
```

## Hypothesis Testing

- Start with a hypothesis (e.g., about the mean ideology in the population)
- Collect data

:::{.fragment}
- Calculate the Z-Score based on null hypothesis:
  - $Z = \frac{\bar{X} - \mu_{0}}{\sigma/\sqrt{n}}$
:::

## Hypothesis Testing 

- Test the null hypothesis using the Z-Score
  - Calculate the p-value 
- P-value: the probability of observing a Z-Score at least as large as that actually observed, if the null hypothesis is true


## Hypothesis Testing

- In R, can use the `t.test()` function

- ...or you can calculate "by hand" and use the `pt()` or `pnorm()` functions


