---
title: "Section 7: Final Meeting"
author: "Sam Frederick"
institute: "Columbia University"
date: "April 25, 2023"
format: 
  revealjs:
    css: [footer.css]
    theme: [default, custom.scss]
    chalkboard: true
    progress: true
    include-after-body: footer.html 
    footer: "Final Section"
knitr: 
  opts_chunk: 
      echo: true
---

```{r setup, include = F, echo = F}
library(tidyverse)
```

## Biases and Errors in Research

```{dot}
digraph G {
    rankdir=LR;
    compound=true;
    Question;
    Theory;
    Hypotheses [group=g1];
    dat [group=g1];
    ht [group=g1]
    {rank=same; dat; msr[group=g2]; sampling[group=g2]; nrb[group=g2]};
    {rank=same; ht; ovb[group= group3]};
Question -> Theory; 
Theory-> Hypotheses;
Hypotheses-> dat;
dat -> sampling;
dat -> msr;
dat -> nrb;
dat -> ht;
ht->ovb;
dat -> ovb;
dat [label = "Data Collection"];
ht [label = "Hypothesis Testing"];
msr[label = "Measurement Error", fillcolor=red, style =filled, fontcolor=white];
sampling[label="Sampling Error", fillcolor=red, style=filled, fontcolor=white];
nrb [label = "Nonresponse Bias", fillcolor = red, style = filled, fontcolor=white];
ovb [label="Confounding Variables", fillcolor=orange, style = filled];
}
```

## Biases and Errors in Research

```{dot}
digraph G {
    rankdir=LR;
    compound=true;
    Question;
    Theory;
    Hypotheses [group=g1];
    dat [group=g1];
    ht [group=g1]
    {rank=same; dat; msr[group=g2]; sampling[group=g2]; nrb[group=g2];rb[group=g2]};
    {rank=same; ht; ovb[group= group3]};
Question -> Theory; 
Theory-> Hypotheses;
Hypotheses-> dat;
dat -> sampling;
dat -> msr;
dat -> nrb;
dat -> ht;
ht->ovb;
dat -> ovb; 
dat-> rb;
dat [label = "Data Collection"];
ht [label = "Hypothesis Testing"];
msr[label = "Measurement Error", fillcolor=red, style =filled, fontcolor=white];
sampling[label="Sampling Error", fillcolor=red, style=filled, fontcolor=white];
nrb [label = "Nonresponse Bias", fillcolor = red, style = filled, fontcolor=white];
rb [label = "Response Bias", fillcolor = red, style = filled, fontcolor=white];
ovb [label="Confounding Variables", fillcolor=orange, style = filled];
}
```

## Response Bias: Social Desirability Bias

:::{.incremental}
- Did you vote in the last election? 
  - Depending on who you're with, you might be more or less likely to answer this question honestly
  
:::


```{r, include = F, echo = F, eval = F}
ces <- haven::read_dta("/Users/samfrederick/Downloads/cumulative_2006-2021.dta")|> 
  filter(year%%2==0) |> 
  mutate(voted_turnout_self = case_when(voted_turnout_self==1~1, 
                                        voted_turnout_self==2~0, 
                                        T~NA_real_), 
         voted_turnout_self_nona =
           case_when(voted_turnout_self==1~1, 
                    voted_turnout_self==0~0, 
                    T~0))
t06 <- read.csv("/Users/samfrederick/Downloads/2006 November General Election - Turnout Rates.csv", skip = 1) |> 
  as_tibble()|> select(X, VEP.Highest.Office) |> 
  filter(X=="United States")|> select(!X)
t08 <- read.csv("/Users/samfrederick/Downloads/2008 November General Election - Turnout Rates.csv", skip = 1) |> 
  as_tibble()|> select(X, VEP.Highest.Office)|> 
  filter(X=="United States")|> select(!X)
t10 <- read.csv("/Users/samfrederick/Downloads/2010 November General Election - Turnout Rates.csv", skip = 1) |> 
  as_tibble()|> select(X, VEP.Highest.Office)|> 
  filter(X=="United States")|> select(!X)
t12 <- read.csv("/Users/samfrederick/Downloads/2012 November General Election v2.0 - Turnout Rates.csv", skip = 1) |> 
  as_tibble()|> select(X, VEP.Highest.Office)|> 
  filter(X=="United States")|> select(!X)
t14 <- read.csv("/Users/samfrederick/Downloads/2014 November General Election - Turnout Rates.csv", skip = 1) |> 
  as_tibble()|> select(X, VEP.Highest.Office)|> 
  filter(X=="United States")|> select(!X)
t16 <- read.csv("/Users/samfrederick/Downloads/2016 November General Election - Turnout Rates.csv", skip = 1) |> 
  as_tibble()|> select(X, VEP.Highest.Office)|> 
  filter(X=="United States")|> select(!X)
t18 <- read.csv("/Users/samfrederick/Downloads/2018 November General Election - Turnout Rates.csv", skip = 1) |> 
  as_tibble()|> select(X, 
                  X2018.Vote.for.Highest.Office.VEP.Turnout.Rate)|> 
  filter(X=="United States")|> select(!X)
t20 <- read.csv("/Users/samfrederick/Downloads/2020 November General Election - Turnout Rates.csv", skip = 1) |> 
  as_tibble()|> select(X, VEP.Turnout.Rate..Highest.Office.)|> 
  filter(X=="United States")|> select(!X)
t22 <- read.csv("/Users/samfrederick/Downloads/2022 November General Election - Turnout Rates.csv", skip = 1) |> 
  as_tibble()|> select(X, Preliminary.Turnout.Rate)|> 
  filter(X=="United States")|> select(!X)
to20 <- read.csv("/Users/samfrederick/Downloads/2020 November General Election - Turnout Rates.csv", skip = 1) |> 
  as_tibble()|> select(X, VEP.Turnout.Rate..Highest.Office.)|> 
  filter(!str_detect(X, "^* Notes")&
           !X%in%c("District of Columbia", "United States"))|> 
  mutate(X = str_remove_all(X, "[^A-Za-z ]"), 
         VEP.Turnout.Rate..Highest.Office. = as.numeric(str_remove(VEP.Turnout.Rate..Highest.Office., "%"))) |> 
  rename("state" = "X")
ces <- haven::read_dta("/Users/samfrederick/Downloads/cumulative_2006-2021.dta")|> 
  filter(year==2020) |> 
  mutate(voted_turnout_self = case_when(voted_turnout_self==1~1, 
                                        voted_turnout_self==2~0, 
                                        T~NA_real_), 
         voted_turnout_self_nona =
           case_when(voted_turnout_self==1~1, 
                    voted_turnout_self==0~0, 
                    T~0))
ces <- ces |> mutate(state_final = labelled::to_factor(state) |> 
                as.character())|> 
  group_by(state_final) |> 
  summarize(voted_turnout_self_nona = 100*mean(voted_turnout_self_nona, na.rm =T))

to20 <- to20 |> 
  left_join(ces, by = c("state" = "state_final"))

to20$state_abb <- state.abb[to22$state%in%state.name]

write_csv(to20, 
          "/Users/samfrederick/scope-and-methods-spring2023/turnout20.csv")

to20 |> mutate(bias = voted_turnout_self_nona - VEP.Turnout.Rate..Highest.Office.) |> ggplot(aes(reorder(state_abb, bias), VEP.Turnout.Rate..Highest.Office.)) + geom_point() + geom_point(aes(reorder(state_abb, bias),voted_turnout_self_nona), color = "blue") + geom_point(aes(reorder(state_abb, bias), bias), color = "red")

turnout <- tibble(year = seq(2006, 2022, 2), 
                  turnout = c(t06$VEP.Highest.Office[1], 
                              t08$VEP.Highest.Office[1], 
                              t10$VEP.Highest.Office[1], 
                              t12$VEP.Highest.Office[1], 
                              t14$VEP.Highest.Office[1], 
                              t16$VEP.Highest.Office[1], 
                              t18$X2018.Vote.for.Highest.Office.VEP.Turnout.Rate[1], 
                              t20$VEP.Turnout.Rate..Highest.Office.[1], 
                              t22$Preliminary.Turnout.Rate[1])) |> 
  mutate(turnout = as.numeric(str_remove(turnout, "%")))

ces <-ces |> group_by(year) |> 
  summarize(voted_turnout_self = 100*mean(voted_turnout_self, 
                                          na.rm = T), 
            voted_turnout_self_nona = 100*mean(voted_turnout_self_nona, na.rm = T))

write_csv(ces, 
          "/Users/samfrederick/scope-and-methods-spring2023/section-7/ces.csv")
write_csv(turnout, 
         "/Users/samfrederick/scope-and-methods-spring2023/section-7/turnout.csv" )

```

## Voter Turnout

:::{.incremental}
- Let's look at some survey results: 

```{r, eval = F}
ces <- read.csv("ces.csv")
```

- What do these results tell us about turnout in elections? 

:::
```{r, echo = F}
ces <- read.csv("/Users/samfrederick/scope-and-methods-spring2023/section-7/ces.csv")
```

## Voter Turnout {auto-animate=true}

- How would we make a scatterplot of turnout from the survey?

```{r, eval = F}
ces 
```

## Voter Turnout {auto-animate=true}

```{r, eval = F}
ces |> 
  ggplot(aes(year, voted_turnout_self_nona))
```

## Voter Turnout {auto-animate=true}

```{r, eval = F}
ces |> 
  ggplot(aes(year, voted_turnout_self_nona))+ 
  geom_point()
```

## Voter Turnout {auto-animate=true}

```{r, eval = F}
ces |> 
  ggplot(aes(year, voted_turnout_self_nona))+ 
  geom_point()+
  ylim(c(0, 100))
```

## Voter Turnout {auto-animate=true}

```{r, eval = F}
ces |> 
  ggplot(aes(year, voted_turnout_self_nona))+ 
  geom_point()+
  ylim(c(0, 100))+
  scale_x_continuous(seq(2006, 2022, 2))
```

## Voter Turnout {auto-animate=true}

```{r, eval = F}
ces |> 
  ggplot(aes(year, voted_turnout_self_nona))+ 
  geom_point()+
  ylim(c(0, 100))+
  scale_x_continuous(seq(2006, 2022, 2))+
  labs(x = "Year", y = "Turnout")
```

## Voter Turnout {auto-animate=true}

```{r, echo = F}
ces |> 
  ggplot(aes(year, voted_turnout_self_nona))+
  geom_point()+
  ylim(c(0, 100))+
  scale_x_continuous(seq(2006, 2022, 2))+
  labs(x = "Year", y = "Turnout")
```

## Voter Turnout

:::{.incremental}
- What about the true turnout? 

```{r, eval = F}
turnout <- read.csv("turnout.csv")
ces <- ces |> 
  left_join(turnout, by = "year")
```

:::

```{r, echo = F}
turnout <- read.csv("/Users/samfrederick/scope-and-methods-spring2023/section-7/turnout.csv")
ces <- ces |> 
  left_join(turnout, by = "year")
```

## Voter Turnout {auto-animate=true}

```{r, eval = F}
#| code-line-numbers: "3"

ces |>
  ggplot(aes(year, voted_turnout_self_nona)) +
  geom_point(aes(color = "Self-Reported Turnout")) +  
  scale_x_continuous(breaks = seq(2006, 2020, 2)) +
  ylim(c(0, 100))+
  labs(x = "Year", y = "Turnout")
```

## Voter Turnout {auto-animate=true}

```{r, eval = F}
ces |>
  ggplot(aes(year, voted_turnout_self_nona)) +
  geom_point(aes(color = "Self-Reported Turnout")) +  
  scale_x_continuous(breaks = seq(2006, 2020, 2)) +
  ylim(c(0, 100))+
  labs(x = "Year", y = "Turnout")
```

## Voter Turnout {auto-animate=true}

```{r, eval =F}
ces |>
  ggplot(aes(year, voted_turnout_self_nona)) +
  geom_point(aes(color = "Self-Reported Turnout")) +  
  scale_x_continuous(breaks = seq(2006, 2020, 2)) +
  ylim(c(0, 100))+
  labs(x = "Year", y = "Turnout")+
  geom_point(aes(year, turnout, color = "True Turnout"))
```

## Voter Turnout {auto-animate=true}

```{r, echo =F}
ces |>
  ggplot(aes(year, voted_turnout_self_nona)) +
  geom_point(aes(color = "Self-Reported Turnout")) +  
  scale_x_continuous(breaks = seq(2006, 2020, 2)) +
  ylim(c(0, 100))+
  labs(x = "Year", y = "Turnout")+
  geom_point(aes(year, turnout, color = "True Turnout"))
```

## Voter Turnout {auto-animate=true}

:::{.incremental}
- What's going on here? 
  - Sampling Error
  - Social Desirability Bias
  - False/Incorrect Memories
  
:::

```{r, echo =F}
ces |>
  ggplot(aes(year, voted_turnout_self_nona)) +
  geom_point(aes(color = "Self-Reported Turnout")) +  
  scale_x_continuous(breaks = seq(2006, 2020, 2)) +
  ylim(c(0, 100))+
  labs(x = "Year", y = "Turnout")+
  geom_point(aes(year, turnout, color = "True Turnout"))
```

## Reducing Social Desirability Bias

:::{.incremental}
- Survey Medium (Face-to-Face, Phone, Online)
- Question Wording
- Anonymized Question Designs

:::

## Anonymized/Secret Question Designs

:::{.incremental}
- List Experiments:
  - Generate a list of items, including the sensitive item you are really interested in 
  - Randomly assign participants to "treatment" and "control"
    - Treatment receives the sensitive item + other items
    - Control receives only other items
  - Ask participants how many of the items apply to them
:::

## Anonymized/Secret Question Designs

:::{.incremental}
- List Experiments:
  - Difference in Means between treatment and control
    - Proportion of people to whom the sensitive item applies
    
:::

## Anonymized/Secret Question Designs

:::{.incremental}
- Things to look out for with List Experiments
  - Ceiling Effects and Floor Effects: 
    - Control items should be a mix of items that should be expected to both apply and not apply to people
    - Violations of this principle removes anonymity
    
:::

## Anonymized/Secret Question Designs

:::{.incremental}
- List Experiments
- Coin flipping, rolling die
  - Tell respondents to flip a coin
  - If outcome heads, give a certain answer
  - If outcome tails, give your true answer
  - Know true proportion who flipped heads (0.5), so difference is true proportion on sensitive answer
  
:::

## Qualitative or Case Study Designs

:::{.incremental}
- Choosing Cases:
  - Think carefully about which cases will help you best test your theory   
  - Select across the range of your independent variable(s)
  - Be aware of the potential limitations of generalization and specific context of cases
  - Consider cases where your theory doesn't hold
    - Does theory need to be refined? 
    - Do these cases suggest limitations/conditions of theory? 
  
:::

## Qualitative or Case Study Designs

:::{.incremental}
- Combining Qualitative and Quantitative
  - Can use mixed methods
    - supplement case study with quantitative analysis
  
:::

## Qualitative or Case Study Designs

:::{.incremental}
- Some Potential Benefits: 
  - In-Depth knowledge about certain cases
  - Better understanding of motivations of political actors
  - Better understanding of context
  - Can bring area-relevant background knowledge of language, culture, etc.
  
:::  

## Qualitative or Case Study Designs

:::{.incremental}
- Some Potential Drawbacks:
  - Representativeness/Generalizability
  - Causal inference can be challenging
    - Need more observations than causes
    - Hard to establish counterfactuals

:::

## Biases and Errors in Research

```{dot}
digraph G {
    rankdir=LR;
    compound=true;
    Question;
    Theory;
    Hypotheses [group=g1];
    dat [group=g1];
    ht [group=g1]
    {rank=same; dat; msr[group=g2]; sampling[group=g2]; nrb[group=g2];rb[group=g2]};
    {rank=same; ht; ovb[group= group3]};
Question -> Theory; 
Theory-> Hypotheses;
Hypotheses-> dat;
dat -> sampling;
dat -> msr;
dat -> nrb;
dat -> ht;
ht->ovb;
dat -> ovb; 
dat-> rb;
dat [label = "Data Collection"];
ht [label = "Hypothesis Testing"];
msr[label = "Measurement Error", fillcolor=red, style =filled, fontcolor=white];
sampling[label="Sampling Error", fillcolor=red, style=filled, fontcolor=white];
nrb [label = "Nonresponse Bias", fillcolor = red, style = filled, fontcolor=white];
rb [label = "Response Bias", fillcolor = red, style = filled, fontcolor=white];
ovb [label="Confounding Variables", fillcolor=orange, style = filled];
}
```



